---
title: "Replenishment Outcome"
date: 2025-03-9
author: "Bruno Alves de Carvalho"
output: pdf_document
---

```{r setup, include=FALSE}
# Set up r-markdown
knitr::opts_chunk$set(echo = FALSE)

# Load clean data
source("Analysis_GC8.R")
```

## Summary
This research aims to identify and assess the influence of political and economic factors on the financial contributions of public donors to the replenishment of The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) from 2001 to 2022. The study draws on publicly available data from reputable sources, including the International Monetary Fund (IMF), the Organisation for Economic Co-operation and Development (OECD), the Center for Global Development (CGD), the Chapel Hill Expert Survey (CHES), and others. It explores the relationship between the financial contributions of key government donors to TGF and their political and economic characteristics, such as political orientation, fiscal and economic outlook, official development assistance (ODA) disbursements, and contributions to other humanitarian organizations. The primary donors analyzed include the United States, Western European countries, and other OECD members, which collectively represent the vast majority of TGF contributions since 2001. This study aims to provide a predictive framework for estimating future replenishment outcomes based on prevailing political and economic conditions, offering TGF a valuable tool to support planning for its eighth grant cycle.

## Research Question
How have political and economic factors influenced the financial contributions of public donors to the replenishment of The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) from 2001 to 2022?

## Introduction
The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) is an international financing organization established in 2002 to combat three of the world’s deadliest epidemics. It operates as a financing mechanism rather than an implementing agency, mobilizing funds from public and private donors and awarding grants to organizations working in low- and middle-income countries. These grants support country-led health programs, developed in collaboration with a broad range of stakeholders, including governments, civil society, affected communities, the private sector, and global health experts.

The Global Fund's reliance on external organizations to implement its grants, known as its partnership model, is both a strength and a vulnerability. On one hand, it allows TGF to leverage the expertise and capacity of its partners. On the other, it makes grant implementation dependent on the viability of these partners. For example, the abrupt axing of the United States Agency for International Development (USAID) by the Trump administration has created financial uncertainty in the humanitarian sector, threatening the execution of TGF grants. Additionally, the current Administration's commitment to global health remains uncertain, adding to concerns about funding sustainability.

To sustain its efforts to end HIV/AIDS, Tuberculosis, and Malaria (HTM), TGF conducts a periodic replenishment, securing new pledges from public and private donors and investing money in three-year grant cycles. Since its inception, it has disbursed $US65 billion, becoming the world’s largest multilateral provider of grants to strengthen health systems. TGF’s financing has contributed to a 63% reduction in combined death rates from HTM, saving 65 million lives (The Global Fund, 2025, pg.6-11).

The eighth replenishment is currently underway, culminating in a high-stakes replenishment conference in November 2025, where donors will confirm their financial commitments. The Global Fund is asking donors for US$18 billion for its eighth grant cycle, running from 2027 to 2029, allowing the Fund to sustain current levels of support to countries. However, in an moment of political and economic uncertainty, where shrinking aid budgets and the de-prioritization of global health funding are reshaping international assistance, the Fund's replenishment is facing stark prospects[^1].

[^1]: Following the termination of their USAID contracts, many organizations are preparing for workforce reductions. UNAIDS is planning to cut 40% of its secretariat staff, including positions in local and regional offices (Ravelo, 2025b). The Stop TB Partnership, headquartered in the same building as The Global Fund in Geneva, has also announced an upcoming downsizing (Ravelo, 2025a). Beyond the health sector, other international organizations are facing similar cuts. The International Organization for Migration (IOM) is set to reduce its workforce by 20%, amounting to 250 job losses at its Geneva headquarters, in addition to laying off 3,000 employees from its U.S. refugee resettlement program (Jerving, 2025). Meanwhile, the UN Refugee Agency (UNHCR) is bracing for up to 6,000 job cuts (Lynch, 2025). As funding reductions continue, further layoffs across international organizations appear inevitable.

In this radical landscape, making predictions may seem speculative. However, it is precisely in times of uncertainty that distinguishing what can be determined with reasonable confidence from what cannot becomes crucial. This research aims to identify key factors that have shaped past TGF replenishment outcomes to inform a predictive framework for estimating future funding trends. Although private sector contributions are briefly considered, the focus remains on public donors, primarily governments, as they provide the majority of The Global Fund’s financing. By doing so, the study seeks to provide The Global Fund with a data-driven tool to navigate a volatile funding environment.

Like many international organizations, The Global Fund relies heavily on the US as its largest donor, with a cumulative contribution of US\$27 billion since 2001 (see Figure 1). The next largest contributors are France, the United Kingdom (UK), Germany, and Japan, having each contributed over US\$5 billion. Historically, TGF's top ten donors have been exclusively high-income members of the Organisation for Economic Co-operation and Development (OECD). These rankings have remained broadly consistent across replenishment cycles, highlighting the Fund's reliance on a stable group of donors for the past two decades.

```{r plot 1, fig.width=7, fig.height=4.1}
# Plot the top 10 biggest donors since 2001 
df_plot <- 
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  filter(donor_type == "public") %>% 
  group_by(donor_name) %>% 
  summarise(sum = sum(pledge_USD, na.rm = T)) %>% 
  slice_max(sum, n = 10)

df_plot %>% 
  ggplot(
    aes(sum, fct_reorder(donor_name, sum))
    ) + 
  geom_bar(
    stat = "identity", 
    fill = grey
    ) +
  labs(
    title = "Figure 1: Top Ten Donors of The Global Fund", 
    x = NULL, y = NULL, 
    caption = source_TGF
    ) + 
  scale_x_continuous(
    breaks = seq(0,30000, 5000), 
    labels = scales::label_currency(suffix = "mil")
    ) +
  plot_frame
```

OECD countries account for over 90% of public sector pledges to The Global Fund, with the United States' share increasing in recent years (see Figure 2 & 3). In 2022, under the Biden Administration, US contributions spiked to 42% of total public sector pledges, reaching US$6 billion. Meanwhile, pledges from non-OECD countries have remained stable since 2001, with no significant increase from BRIC members such as China, Brazil, or India. This suggests that these countries are unlikely to offset expected funding cuts from the US, France, UK, and Germany, raising concerns about potential funding gaps.

```{r plot 2 & 3}
# Plot the share (%) of financial contributions by country from 2001 to 2023
p_relative_contributions <- 
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), pledge_USD, fill = oecd_status)) + 
  geom_bar(stat = "identity", position = "fill") +
  labs(
    title = "Figure 2 & 3: Perentage and Dollar Share of Financial Contributions",
    x = NULL, y = NULL,
    caption = source_TGF
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c(red, yellow, blue)) +
  plot_frame +
  theme(legend.position = "top",
        legend.title = element_blank()
        ) +
  guides(
    fill= guide_legend(reverse = TRUE)
    )

# Plot the absolute financial contribution by country from 2001 to 2023
p_absolute_contributions <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), pledge_USD, fill = oecd_status)) + 
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    x = NULL, y = NULL
  ) +
  scale_y_continuous(labels = scales::label_currency(suffix = "mil")) +
  scale_fill_manual(values = c(red, yellow, blue)) +
  plot_frame 

# Combine both plots
p_relative_contributions + p_absolute_contributions
```

Despite its reliance on a small group of major donors, TGF has expanded its donor base since 2016, particularly engaging the private sector more effectively. The share of private sector pledges increased from 3% in 2001 to 8% in 2022, although it has remained stagnant since 2019 (see Figure 4). This growth reflects TGF's ability to leverage new technologies and private sector expertise to develop smarter solutions for combating HTM, attracting large donations from foundations like the Bill & Melinda Gates Foundation and (RED).

Additionally, the number of public sector donors was initially high and met with excitement when the Fund launched in 2001 but stabilized at 30 after the adoption of the Replenishment model in 2005 (see Figure 5). It began rising again in 2016, nearly doubled by 2019, and reached almost 50 in 2022. Most new donors are low to middle-income countries, reflecting a positive shift toward greater self-reliance among recipient nations. However, despite this progress, the financial contributions from these new donors remain relatively small. As a result, they cannot be relied on as a hedge against a potential funding drop from the US and other European countries, especially since these countries' budgets would also be affected by a global reduction in international aid.

Furthermore, the expansion of the donor base has coincided with a decline in the median financial contribution per country (see Figure 6). This suggests that simply increasing the number of donors is not enough to boost overall funding levels. To sustain and enhance support for recipient countries, it is equally important to maintain high individual country contributions. Nevertheless, a more diverse donor base still holds value as a hedge against funding risks—a direction that many organizations and recipient countries will need to explore as the US limits its involvement in international aid.

```{r plot 4 & 5, fig.width=7, fig.height=4.5}
# Plot the share of private sector donations
p_private_relative_contributions <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  group_by(year) %>% 
  mutate(prct = pledge_USD / sum(pledge_USD)) %>% 
  filter(donor_type == "private") %>% 
  ggplot(aes(as.factor(year), prct)) + 
  geom_bar(stat = "identity", fill = grey) +
  labs(
    title = "Figure 4 & 5: Percentage of Private Sector Contributions (left) and Number of Public\nSector Donors (right)",
    x = NULL, y = NULL,
    caption = source_TGF
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  plot_frame

# Plot the number of government donors per grant cycle
p_num_donors <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  filter(donor_type == "public") %>% 
  group_by(year) %>% 
  count() %>% 
  ggplot(aes(as.factor(year), n)) + 
  geom_bar(stat = "identity", fill = grey) +
  labs(
    x = NULL, y = NULL
  ) +
  scale_y_continuous(breaks = c(0, 30, 40, 50, 60)) +
  plot_frame 

# Combine both plots
p_private_relative_contributions + p_num_donors 
```

```{r plot 6, fig.width=7, fig.height=5}
# Plot the median financial country contribution from 2001 to 2023
list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), log(pledge_USD))) + 
  geom_jitter(color = grey, alpha = 0.5) + 
  geom_boxplot() +
  labs(
    title = "Figure 6: Median Contribution per Public Sector Donor (log-scale)",
    x = NULL, y = NULL,
    caption = source_TGF
  ) +
  plot_frame
```

Overall, The Global Fund has benefited from consistent growth in funding, enabling it to improve health outcomes in developing countries and earn international recognition. While the success of its grants has undoubtedly played a role in securing continued support, other factors may have also influenced public donor contributions. This question is particularly relevant in an era when financing appears to be driven less by the measurable impact of international aid and more by the alignment of that aid with donor countries’ national interests. In other words, development practitioners must look beyond the effectiveness of their programs and examine the economic and political conditions shaping public donor contributions.

Financial contributions can be understood through the lens of supply and demand. The demand for funding is represented by current country needs, as well as trends in the incidence and mortality of HIV/AIDS, tuberculosis, and malaria (HTM). Based on these factors, The Global Fund, in collaboration with academic experts, has estimated that US$18 billion is needed to stay on track to end HTM by 2030. On the supply side, a country’s ability and willingness to contribute depends on several economic and political considerations.

One way to assess this is through trends in Official Development Assistance (ODA), which reflect the total volume of aid disbursed by donor countries. Higher ODA levels may indicate a greater likelihood of increased contributions to The Global Fund. Similarly, if other international aid organizations receive greater public sector contributions, The Global Fund may also benefit, given that these organizations operate within the same sector and under similar funding conditions.

More fundamentally, the extent to which funding supply matches demand depends on the broader political and economic context of donor countries. For example, a government experiencing strong economic growth and fiscal stability may be better positioned to contribute than one facing recession and high public debt. Political ideology may also play a role: left-leaning governments, which typically emphasize universal healthcare and poverty reduction, may be more inclined to support humanitarian initiatives, whereas right-leaning governments may prioritize domestic spending and military investments. Additionally, electoral cycles could impact funding commitments. During an election year, governments may be preoccupied with domestic politics, making it harder for The Global Fund to make its investment case heard.

However, some governments may still maintain or even increase contributions due to pre-existing commitments or a desire to uphold international credibility, regardless of political and economic conditions. The generalizations outlined above may not apply universally. These considerations form the basis of six hypotheses that this research will test using publicly available data from multiple reputable sources. Namely, we collected data from the International Monetary Fund (IMF), the Organisation for Economic Co-operation and Development (OECD), the Center for Global Development (CGD), the Chapel Hill Expert Survey (CHES), the Party Government in Europe Database (PAGED), the ParlGov project, and the Manifesto Project database (MP).

Having analysed the historical replenishment outcomes of The Global Fund and outlined our hypotheses, we will now dig deeper into the data to test our  assumptions, identify patterns and trends in donor behavior, and attempt to model these behaviors to create a predictive framework for estimating future funding outcomes. Following that, we will discuss the implications of the findings and offer recommendations for the future sustainability of The Global Fund, particularly in light of the evolving political and economic challenges.

## Hypotheses
1. If the financial contributions from government donors to other global health organizations, such as GAVI, and multilateral development banks, like the IDA, decline, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because TGF tends to align its funding expectations with global trends in public health financing.
2. If the Official Development Assistance (ODA) disbursements of government donors decline, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because TGF's funding is indirectly influenced by overall humanitarian aid trends.
3. If the fiscal outlook of government donors, measured *inter lia* by rising public debt and deteriorating fiscal balances, worsens, then the financial contributions from government donors to The Global Fund (TGF) will decrease, because weaker fiscal health limits governments' ability to make discretionary international contributions.
4. If the macroeconomic outlook of government donors, measured *inter lia* by GDP growth and unemployment, worsens, then the financial contributions from government donors to The Global Fund (TGF) will decrease, because weaker economic growth limits govertnments' ability to make discretionary international contributions.
5. If the ideological placement of government donors, measured by their political and / or economic orientation, leans to the right, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because right-wing governments tend to prioritize fiscal discipline and are less ideologically aligned with humanitarian spending.
6. If the replenishment year coincides with the re-election year of donor governments, then their financial contributions to The Global Fund (TGF) will be smaller, because electoral priorities and domestic political considerations may reduce focus on international aid commitments and make it more challenging for TGF to secure funding.

## Hypotheses Testing
Ongoing

## Recommendations
Not Started

## Sources
Döring, H. and Manow, P. (2024) ParlGov 2024 Release. Available at: https://doi.org/10.7910/DVN/2VZ5ZC (Accessed: 22/03/2025).

Hellström, J., Bergman, T., Lindahl, J. and Gustafsson, E.K. (2025) The Representative Democracy Data Archive (REPDEM) – Basic dataset, Version 2025.03. Available at: https://repdem.org (Accessed: 22/03/2025).

International Monetary Fund (IMF) (2024) Fiscal Monitor (FM) Dataset. Available at: https://betadata.imf.org/en/Data-Explorer?datasetUrn=IMF.FAD:FM(5.0.0) (Accessed: 22/02/2025).

International Monetary Fund (IMF) (2024) World Economic Outlook (WEO) Database, October 2024. Available at: https://www.imf.org/en/Publications/WEO/weo-database/2024/October (Accessed: 22/03/2025).

Jerving, S. (2025) ‘Scoop: UN migration agency cuts more jobs, shutters lifesaving programs’, Devex, 20 March. Available at: https://www.devex.com/news/memo-lays-out-plan-to-replace-usaid-with-new-humanitarian-agency-109685 (Accessed: 22 March 2025).

Jolly, S., Bakker, R., Hooghe, L., Marks, G., Polk, J., Rovny, J., Steenbergen, M. and Vachudova, M.A. (2022) ‘Chapel Hill Expert Survey Trend File, 1999-2019’, Electoral Studies, 75 (February). Available at: https://doi.org/10.1016/j.electstud.2021.102420 (Accessed: 22/03/2025).

Lehmann, P., Franzmann, S., Al-Gaddooa, D., Burst, T., Ivanusch, C., Regel, S., Riethmüller, F., Volkens, A., Weßels, B. and Zehnter, L. (2024) The Manifesto Data Collection. Manifesto Project (MRG/CMP/MARPOR). Version 2024a. Berlin: Wissenschaftszentrum Berlin für Sozialforschung (WZB) / Göttingen: Institut für Demokratieforschung (IfDem). Available at: https://doi.org/10.25522/manifesto.mpds.2024a (Accessed: 22/03/2025).

Lynch, C. (2025) ‘Exclusive: UN Refugee Agency braces for thousands of job cuts’, Devex, 20 March. Available at: https://www.devex.com/news/exclusive-un-refugee-agency-braces-for-thousands-of-job-cuts-109693 (Accessed: 22 March 2025).

Madan Keller, J., Landers, C., Martinez, N. and Eldridge, R. (2024) ‘The 2024–2025 replenishment traffic jam’, Center for Global Development, 16 October. Available at: https://www.cgdev.org/media/2024-2025-replenishment-traffic-jam (Accessed: 22/03/2025).

Ravelo, J.L. (2025a) ‘Stop TB Partnership looks at downsizing after USAID cuts’, Devex, 4 March. Available at: https://www.devex.com/news/stop-tb-partnership-looks-at-downsizing-after-usaid-cuts-109544 (Accessed: 22 March 2025).

Ravelo, J.L. (2025b) ‘Exclusive: UNAIDS staff anxious as restructuring looms amid aid cuts’, Devex, 14 March. Available at: https://www.devex.com/news/exclusive-unaids-staff-anxious-as-restructuring-looms-amid-aid-cuts-109638 (Accessed: 22 March 2025).

Rovny, J., Bakker, R., Hooghe, L., Jolly, S., Marks, G., Polk, J., Steenbergen, M. and Vachudova, M.A. ‘25 Years of Political Party Positions in Europe: The Chapel Hill Expert Survey, 1999-2024’, working paper.

The Global Fund (2025) The Global Fund Investment Case. Available at: https://www.theglobalfund.org/en/investment-case/ (Accessed: 22/03/2025).
