---
title: "Replenishing Hope: Political and Economic Drivers of Public Sector Support for The Global Fund"
date: 2025-03-23
author: "Bruno Alves de Carvalho"
output: pdf_document
---

```{r setup, include=FALSE}
# Set up r-markdown
knitr::opts_chunk$set(echo = FALSE)

# Load clean data
source("Analysis_GC8.R")
```

## Summary
This research aims to identify and assess the influence of political and economic factors on the financial contributions of public donors to the replenishment of The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) from 2001 to 2022. The study draws on publicly available data from reputable sources, including the International Monetary Fund (IMF), the Organisation for Economic Co-operation and Development (OECD), the Center for Global Development (CGD), the Chapel Hill Expert Survey (CHES), and others. It explores the relationship between the financial contributions of key government donors to TGF and their political and economic characteristics, such as political orientation, fiscal and economic outlook, official development assistance (ODA) disbursements, and contributions to other humanitarian organizations. The primary donors analyzed include the United States, Western European countries, and other OECD members, which collectively represent the vast majority of TGF contributions since 2001. This study aims to provide a predictive framework for estimating future replenishment outcomes based on prevailing political and economic conditions, offering TGF a valuable tool to support planning for its eighth grant cycle.

## Research Question
How have political and economic factors influenced the financial contributions of public donors to the replenishment of The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) from 2001 to 2022?

## Introduction
The Global Fund to Fight AIDS, Tuberculosis, and Malaria (TGF) is an international financing organization established in 2002 to combat three of the world’s deadliest epidemics. It operates as a financing mechanism rather than an implementing agency, mobilizing funds from public and private donors and awarding grants to organizations working in low- and middle-income countries. These grants support country-led health programs, developed in collaboration with a broad range of stakeholders, including governments, civil society, affected communities, the private sector, and global health experts.

The Global Fund's reliance on external organizations to implement its grants is know as a partnership model. This model allows TGF to leverage the expertise and capacity of its partners, though it makes grant implementation dependent on the viability of these partners. For example, the recent axing of the United States Agency for International Development (USAID) by the Trump administration has created financial uncertainty in the humanitarian sector that could potentially threaten the execution of some TGF grants[^1].

[^1]: Following the termination of their USAID contracts, many organizations that TGF partners with have had to reduce their scope of work and workforce. For example, UNAIDS is planning to cut 40% of its secretariat staff, including positions in local and regional offices (Ravelo, 2025b). Similarly, The Stop TB Partnership, headquartered in the same building as The Global Fund in Geneva, has also announced an upcoming downsizing (Ravelo, 2025a). Beyond The Global Fund's partnership, other international organizations are facing similar cuts. The International Organization for Migration (IOM) is set to reduce its workforce by 20%, amounting to 250 job losses at its Geneva headquarters, in addition to laying off 3'000 employees from its U.S. refugee resettlement program (Jerving, 2025a). Meanwhile, the UN Refugee Agency (UNHCR) is bracing for up to 6'000 job cuts (Lynch, 2025a). As funding reductions continue, further layoffs across international organizations appear inevitable.

To sustain its efforts to end HIV/AIDS, Tuberculosis, and Malaria (HTM), TGF conducts a periodic replenishment, securing new pledges from public and private donors and investing money in three-year grant cycles. Since its inception, it has disbursed $US65 billion, becoming the world’s largest multilateral provider of grants to strengthen health systems. TGF’s financing has contributed to a 63% reduction in combined death rates from HTM, saving 65 million lives (The Global Fund, 2025, pg.6-11).

The eighth replenishment is currently underway, culminating in a high-stakes replenishment conference in November 2025, where donors will confirm their financial commitments. The Global Fund is asking donors for US$18 billion for its eighth grant cycle, running from 2027 to 2029, allowing the Fund to sustain current levels of support to countries. However, in an moment of political and economic uncertainty, where shrinking aid budgets and the de-prioritization of global health funding are reshaping international assistance, the Fund is facing a challenging replenishment[^2].

[^2]: As the US undergoes a major overhaul of its international aid apparatus, key European donors are also scaling back their foreign assistance. France, facing fiscal pressures, has reduced its aid budget by 35%. The UK plans to cut its foreign aid allocation from 0.5% to 0.3% of gross national income by 2027, prioritizing defense spending. Additionally, Belgium, the Netherlands, and Switzerland have announced similar reductions in aid budgets (Galvin, 2025). Germany is likely to follow suit. However, there is some optimism. Reports on the US government's reshaped international aid strategy seem to indicate that global health remains a priority. A memo, reportedly drafted by officials from the Trump Administration, suggests that the success of global health initiatives should be measured by factors such as lives saved and the containment of infectious disease outbreaks, two areas where The Global Fund has demonstrated particular effectiveness (Toosi and Lippman, 2025; Jerving, 2025b).

In this radical landscape, making predictions may seem speculative. However, it is precisely in times of uncertainty that distinguishing what can be determined with reasonable confidence from what cannot becomes crucial. This research aims to identify key factors that have shaped past TGF replenishment outcomes to inform a predictive framework for estimating future funding trends. Although private sector contributions are briefly considered, the focus remains on public donors, primarily governments, as they provide the majority of The Global Fund’s financing. By doing so, the study seeks to provide The Global Fund with a data-driven tool to navigate a volatile funding environment.

Since its launch in 2001, public donor pledges to The Global Fund have varied significantly. The vast majority of donor countries contribute between US\$1 million and US\$200 million, while only a few have pledged over US\$1 billion (see Figure 1). High-value pledges are rare but substantial, with the United States being the only country to have made multi-billion-dollar contributions.

Like many international organizations, TGF relies heavily on the US as its largest donor, with a cumulative contribution of US\$27 billion since 2001 (see Figure 2). This dominance underscores the US’s strong commitment to multilateralism over the past two decades. Beyond the US, the next largest contributors are France, the United Kingdom (UK), Germany, and Japan, having each contributed over US\$5 billion. Historically, TGF's top ten donors have been exclusively high-income members of the Organisation for Economic Co-operation and Development (OECD). These rankings have remained broadly consistent across replenishment cycles, highlighting the Fund's reliance on a stable group of donors for the past two decades.

OECD countries account for over 90% of public sector pledges to The Global Fund, with the United States' share increasing in recent years (see Figure 2 & 3). In 2022, under the Biden Administration, US contributions spiked to 42% of total public sector pledges, reaching US$6 billion. Meanwhile, pledges from non-OECD countries have remained stable since 2001, with no significant increase from BRIC members such as China, Brazil, or India. This suggests that BRIC countries are unlikely to offset expected funding cuts from the US and key European donors, raising concerns about potential funding gaps. More importantly, the lack of support from BRIC nations may signal a failure to effectively engage with these donors, though it is known that China, in particular, prefers bilateral aid over multilateral contributions.

```{r plot 1 & 2, fig.height=4.25}
# Plot the distribution of pledges
p_pledge_distribution <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  filter(donor_type == "public" & pledge_USD > 0) %>% 
  ggplot(aes(pledge_USD)) + 
  geom_histogram(bins = 30) +
  labs (
    title = "Figure 1 & 2: Pledge Distribution and Top Ten Donors of The Global Fund\nfrom 2001 to 2022", 
    x = NULL, 
    y = NULL,
    caption = source_TGF
    ) + 
  scale_x_continuous(
    breaks = c(0, 2000, 4000, 6000),
    labels = scales::label_currency(suffix = "mil")
    ) + 
  plot_frame

# Plot the top 10 biggest donors since 2001 
df_plot <- 
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  filter(donor_type == "public") %>% 
  group_by(donor_name) %>% 
  summarise(sum = sum(pledge_USD, na.rm = T)) %>% 
  slice_max(sum, n = 10)

p_top_donors <-
  df_plot %>% 
  ggplot(
    aes(sum, fct_reorder(donor_name, sum))
    ) + 
  geom_bar(
    stat = "identity", 
    fill = grey
    ) +
  labs(
    x = NULL, 
    y = NULL
    ) + 
  scale_x_continuous(
    breaks = c(5000, 15000, 25000), 
    labels = scales::label_currency(suffix = "mil")
    ) +
  plot_frame

# Combine plots
p_pledge_distribution + p_top_donors
```

```{r plot 3 & 4}
# Plot the share (%) of financial contributions by country from 2001 to 2023
p_relative_contributions <- 
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), pledge_USD, fill = oecd_status)) + 
  geom_bar(stat = "identity", position = "fill") +
  labs(
    title = "Figure 3 & 4: Perentage and Dollar Share of Financial Contributions",
    x = NULL, y = NULL,
    caption = source_TGF
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c(red, yellow, blue)) +
  plot_frame +
  theme(legend.position = "top",
        legend.title = element_blank()
        ) +
  guides(
    fill= guide_legend(reverse = TRUE)
    )

# Plot the absolute financial contribution by country from 2001 to 2023
p_absolute_contributions <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), pledge_USD, fill = oecd_status)) + 
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    x = NULL, y = NULL
  ) +
  scale_y_continuous(labels = scales::label_currency(suffix = "mil")) +
  scale_fill_manual(values = c(red, yellow, blue)) +
  plot_frame 

# Combine both plots
p_relative_contributions + p_absolute_contributions
```

Despite its reliance on a small group of major donors, TGF has successfully expanded its donor base since 2016, notably through private sector engagement and new public sector contributors. The share of private sector pledges increased from 3% in 2001 to 8% in 2022, although it has remained stagnant since 2019 (see Figure 4). This growth reflects TGF's ability to leverage new technologies and private sector expertise to develop smarter solutions for combating HTM, attracting large donations from foundations like the Bill & Melinda Gates Foundation and (RED). Likewise, the number of public sector donors increased from 28 in 2013 to 36 in 2016, reaching 58 in 2019 and 49 in 2022 (see Figure 5). This is higher than the initial group of public donors in 2001, who benefited from the excitement surrounding the creation of this innovative financing mechanism. Most new donors are low to middle-income countries, reflecting a positive shift toward greater self-financing among recipient nations. 

However, despite this broader participation, contributions from new donors remain relatively small, suggesting their pledges are more intended to be symbolic commitments rather than financially transformative. This means they cannot compensate for potential reductions from major donors like the US and key European countries. Additionally, while the number of donors has increased, the median pledge per country has declined, from US\$46.2 million in 2013 to US\$10 million in 2022 (see Figure 6). This underscores a critical point: increasing the number of donors alone is not enough to sustain overall funding levels. To build a more resilient financial model, TGF must attract a greater share of mid-sized donors, balancing its funding base and reducing dependence on a single dominant donor[^3]. 

[^3]: Exploring alternative funding sources beyond donors could also serve as a hedge against funding risks. This is particularly relevant for The Global Fund, which relies entirely on donor funding to finance its operations and investments. Such an approach might offer a more attractive hedge than attempting to expand the donor base, as the latter could become increasingly difficult in a world where globalization is retreating. There are already signs that the US may reduce funding to international organizations that receive financial support from China and other countries competing with America's global influence (Lynch, 2025b).

```{r plot 5 & 6, fig.width=7, fig.height=4.5}
# Plot the share of private sector donations
p_private_relative_contributions <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  group_by(year) %>% 
  mutate(prct = pledge_USD / sum(pledge_USD)) %>% 
  filter(donor_type == "private") %>% 
  ggplot(aes(as.factor(year), prct)) + 
  geom_bar(stat = "identity", fill = grey) +
  labs(
    title = "Figure 5 & 6: Percentage of Private Sector Contributions (left) and Number of Public\nSector Donors (right)",
    x = NULL, y = NULL,
    caption = source_TGF
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  plot_frame

# Plot the number of government donors per grant cycle
p_num_donors <-
  list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% 
  filter(donor_type == "public") %>% 
  group_by(year) %>% 
  count() %>% 
  ggplot(aes(as.factor(year), n)) + 
  geom_bar(stat = "identity", fill = grey) +
  labs(
    x = NULL, y = NULL
  ) +
  scale_y_continuous(breaks = c(0, 30, 40, 50, 60)) +
  plot_frame 

# Combine both plots
p_private_relative_contributions + p_num_donors 
```

```{r plot 7, fig.width=7, fig.height=5}
# Plot the median financial country contribution from 2001 to 2023
list_data %>% 
  filter(source == "FULL_FACTORS") %>% 
  pluck(2,1) %>% filter(donor_type == "public") %>% 
  ggplot(aes(as.factor(year), log(pledge_USD))) + 
  geom_jitter(color = grey, alpha = 0.5) + 
  geom_boxplot() +
  labs(
    title = "Figure 7: Median Contribution per Public Sector Donor (log-scale)",
    x = NULL, y = NULL,
    caption = source_TGF
  ) +
  plot_frame
```

Overall, the skewed distribution of pledges reveals deep structural realities about international funding: economic disparities, donor dominance, and long-term sustainability risks. Even though TGF has made progress in diversifying its donor pool, the financial weight remains highly concentrated among a few key players. These issues notwithstanding, the Global Fund has benefited from consistent growth in public sector funding, from US\$4.7 billion in 2001 to US\$14 billion in 2022, enabling it to improve health outcomes in developing countries and earn international recognition. While the success of its grants has undoubtedly played a role in securing continued support, other factors may have also influenced public donor contributions. This question is particularly relevant in an era when financing appears to be driven less by the measurable impact of international aid and more by the alignment of that aid with donor countries’ national interests. In other words, we must look beyond the effectiveness of aid programs and examine the economic and political conditions shaping public donor contributions.

The extent of public sector pledges to The Global Fund can be understood as a function of supply and demand. The demand for funding is represented by current country needs, as well as trends in the incidence and mortality of HIV/AIDS, tuberculosis, and malaria (HTM). Based on these factors, The Global Fund, in collaboration with academic experts, has estimated that US$18 billion is needed between 2027 and 2029 to stay on track to end HTM by 2030. On the supply side, a country’s ability and willingness to contribute money to the partnership can be inferred from the overall pool of funds allocated for foreign aid. One way to assess this is through trends in Official Development Assistance (ODA), which reflect the total volume of aid disbursed by donor countries. Higher ODA levels may indicate a greater likelihood of increased contributions to The Global Fund. Similarly, if other international aid organizations receive greater public sector contributions, The Global Fund may also benefit, given that these organizations operate within the same sector and under similar funding conditions.

More fundamentally, the extent to which the supply of funds matches demand depends on the broader political and economic context of donor countries. For example, a government experiencing strong economic growth and fiscal stability may be better positioned to contribute than one facing recession and high public debt. Political ideology may also play a role: left-leaning governments, which typically emphasize universal healthcare and poverty reduction, may be more inclined to support humanitarian initiatives, whereas right-leaning governments may prioritize domestic spending and military investments. Additionally, electoral cycles could impact funding commitments. During an election year, governments may be preoccupied with domestic politics, making it harder for The Global Fund to make its investment case heard.

However, some governments may still maintain or even increase contributions due to pre-existing commitments or a desire to uphold international credibility, regardless of political and economic conditions. The generalizations outlined above may not apply universally. These considerations form the basis of six hypotheses that this research will test using publicly available data from multiple reputable sources, namely: the International Monetary Fund (IMF), the OECD, the Center for Global Development (CGD), the Chapel Hill Expert Survey (CHES), the Party Government in Europe Database (PAGED), the ParlGov project, and the Manifesto Project database (MP).

Having analysed the historical replenishment outcomes of The Global Fund and outlined our hypotheses, we will now dig deeper into the data to test our  assumptions, identify patterns and trends in donor behavior, and attempt to model these behaviors to create a predictive framework for estimating future funding outcomes. Following that, we will discuss the implications of the findings and offer recommendations for the future sustainability of The Global Fund, particularly in light of the evolving political and economic challenges.

## Hypotheses
1. If the financial contributions from government donors to other global health organizations, such as GAVI, and multilateral development banks, like the IDA, decline, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because TGF tends to align its funding expectations with global trends in public health financing.
2. If the Official Development Assistance (ODA) disbursements of government donors decline, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because TGF's funding is indirectly influenced by overall humanitarian aid trends.
3. If the fiscal outlook of government donors, measured *inter lia* by rising public debt and deteriorating fiscal balances, worsens, then the financial contributions from government donors to The Global Fund (TGF) will decrease, because weaker fiscal health limits governments' ability to make discretionary international contributions.
4. If the macroeconomic outlook of government donors, measured *inter lia* by GDP growth and unemployment, worsens, then the financial contributions from government donors to The Global Fund (TGF) will decrease, because weaker economic growth limits govertnments' ability to make discretionary international contributions.
5. If the ideological placement of government donors, measured by their political and / or economic orientation, leans to the right, then the financial contributions from public donors to The Global Fund (TGF) will decrease, because right-wing governments tend to prioritize fiscal discipline and are less ideologically aligned with humanitarian spending.
6. If the replenishment year coincides with the re-election year of donor governments, then their financial contributions to The Global Fund (TGF) will be smaller, because electoral priorities and domestic political considerations may reduce focus on international aid commitments and make it more challenging for TGF to secure funding.

## Hypotheses Testing
In an article from 2024, researchers from the CDG showed how multiple organizations were seeking funds at a time of constrained donor funding. Coming at the end of this traffic jam, can we use the replenishment results of other organizations to estimate what TGF might receive? We will try to answer this question.

## Recommendations
Not Started

## Sources
Döring, H. and Manow, P. (2024) ParlGov 2024 Release. Available at: https://doi.org/10.7910/DVN/2VZ5ZC (Accessed: 22/03/2025).

Galvin, G. (2025) ‘“Utterly devastating”: Global health groups left reeling as European countries slash foreign aid’, Euronews, 7 March. Available at: https://www.euronews.com/health/2025/03/07/utterly-devastating-global-health-groups-left-reeling-as-european-countries-slash-foreign- (Accessed: 23 March 2025).

Hellström, J., Bergman, T., Lindahl, J. and Gustafsson, E.K. (2025) The Representative Democracy Data Archive (REPDEM) – Basic dataset, Version 2025.03. Available at: https://repdem.org (Accessed: 22/03/2025).

International Monetary Fund (IMF) (2024) Fiscal Monitor (FM) Dataset. Available at: https://betadata.imf.org/en/Data-Explorer?datasetUrn=IMF.FAD:FM(5.0.0) (Accessed: 22/02/2025).

International Monetary Fund (IMF) (2024) World Economic Outlook (WEO) Database, October 2024. Available at: https://www.imf.org/en/Publications/WEO/weo-database/2024/October (Accessed: 22/03/2025).

Jerving, S. (2025a) ‘Scoop: UN migration agency cuts more jobs, shutters lifesaving programs’, Devex, 20 March. Available at: https://www.devex.com/news/memo-lays-out-plan-to-replace-usaid-with-new-humanitarian-agency-109685 (Accessed: 22 March 2025).

Jerving, S. (2025b) ‘Memo lays out plan to replace USAID with new humanitarian agency’, Devex, 20 March. Available at: https://www.devex.com/news/memo-lays-out-plan-to-replace-usaid-with-new-humanitarian-agency-109685 (Accessed: 23 March 2025).

Jolly, S., Bakker, R., Hooghe, L., Marks, G., Polk, J., Rovny, J., Steenbergen, M. and Vachudova, M.A. (2022) ‘Chapel Hill Expert Survey Trend File, 1999-2019’, Electoral Studies, 75 (February). Available at: https://doi.org/10.1016/j.electstud.2021.102420 (Accessed: 22/03/2025).

Lehmann, P., Franzmann, S., Al-Gaddooa, D., Burst, T., Ivanusch, C., Regel, S., Riethmüller, F., Volkens, A., Weßels, B. and Zehnter, L. (2024) The Manifesto Data Collection. Manifesto Project (MRG/CMP/MARPOR). Version 2024a. Berlin: Wissenschaftszentrum Berlin für Sozialforschung (WZB) / Göttingen: Institut für Demokratieforschung (IfDem). Available at: https://doi.org/10.25522/manifesto.mpds.2024a (Accessed: 22/03/2025).

Lynch, C. (2025a) ‘Exclusive: UN Refugee Agency braces for thousands of job cuts’, Devex, 20 March. Available at: https://www.devex.com/news/exclusive-un-refugee-agency-braces-for-thousands-of-job-cuts-109693 (Accessed: 22 March 2025).

Lynch, C. (2025b) ‘Trump’s new ‘loyalty test’ for UN and aid groups’, Devex, 12 March. Available at: https://www.devex.com/news/trump-s-new-loyalty-test-for-un-and-aid-groups-109635 (Accessed: 23 March 2025).

Madan Keller, J., Landers, C., Martinez, N. and Eldridge, R. (2024) ‘The 2024–2025 replenishment traffic jam’, Center for Global Development, 16 October. Available at: https://www.cgdev.org/media/2024-2025-replenishment-traffic-jam (Accessed: 22/03/2025).

Ravelo, J.L. (2025a) ‘Stop TB Partnership looks at downsizing after USAID cuts’, Devex, 4 March. Available at: https://www.devex.com/news/stop-tb-partnership-looks-at-downsizing-after-usaid-cuts-109544 (Accessed: 22 March 2025).

Ravelo, J.L. (2025b) ‘Exclusive: UNAIDS staff anxious as restructuring looms amid aid cuts’, Devex, 14 March. Available at: https://www.devex.com/news/exclusive-unaids-staff-anxious-as-restructuring-looms-amid-aid-cuts-109638 (Accessed: 22 March 2025).

Rovny, J., Bakker, R., Hooghe, L., Jolly, S., Marks, G., Polk, J., Steenbergen, M. and Vachudova, M.A. ‘25 Years of Political Party Positions in Europe: The Chapel Hill Expert Survey, 1999-2024’, working paper.

The Global Fund (2025) The Global Fund Investment Case. Available at: https://www.theglobalfund.org/en/investment-case/ (Accessed: 22/03/2025).

Toosi, N. and Lippman, D. (2025) ‘Trump aides circulate plan for complete revamp of foreign aid programs’, Politico, 19 March. Available at: https://www.politico.com/news/2025/03/19/trump-aides-circulate-plan-for-complete-revamp-of-foreign-aid-programs-00238862 (Accessed: 23 March 2025).
